Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-04-16 paper/abstract.tex
second pass over intro/related
Wordsmithing abstract

% Program logics and semantics tell us that in order to derive
% \begin{math}
%   ((S_1\SEMI S_2), \sigma_0) \Downarrow \sigma_2,
% \end{math}
% we derive 
% \begin{math}
%   (S_1, \sigma_0) \Downarrow \sigma_1
% \end{math}
% and then
% \begin{math}
%   (S_2, \sigma_1) \Downarrow \sigma_2.
% \end{math}
Program logics and semantics tell us that when executing ((S1; S2), state0),
we execute (S1, state0) to arrive at state1, then execute (S2, state1) to
arrive at the final state2.  This is a delightfully simple story that can be
explained to children.  It is also a lie.
%
Processors execute instructions out of order, due to pipelines and caches.
Compilers reorder programs even more dramatically.  All of this reordering is
meant to be unobservable in single-threaded code.  In multi-threaded code,
however, all bets are off.  A formal attempt to understand the resulting mess
is known as a ``relaxed memory model.''  The relaxed memory models that have
been proposed to date either fail to address sequential composition, or
overly restrict processors and compilers.

To support sequential composition, we propose adding families of predicate
transformers to the existing model of ``Pomsets with Preconditions,'' which
already supports parallel composition.  When composing (S1;S2), the predicate
transformer used to validate the precondition of an event in S2 is chosen
based on the semantic dependencies from S1 into this event.  Our model
retains the good properties of the prior work, including efficient
implementation on Arm8, support for compiler optimizations, support for
logics that prove the absence of thin-air behaviors, and a local data race
freedom theorem.

>>>>

% Program logics and semantics tell us that in order to derive
% \begin{math}
%   ((S_1\SEMI S_2), \sigma_0) \Downarrow \sigma_2,
% \end{math}
% we derive 
% \begin{math}
%   (S_1, \sigma_0) \Downarrow \sigma_1
% \end{math}
% and then
% \begin{math}
%   (S_2, \sigma_1) \Downarrow \sigma_2.
% \end{math}
Program logics and semantics tell us that when executing $(S_1; S_2)$ starting in state $s_0$,
we execute $S_1$ in $s_0$ to arrive at $s_1$, then execute $S_2$ in $s_1$ to
arrive at the final state $s_2$.  This is, of course, an abstraction.  Processors
execute instructions out of order, due to pipelines and caches, and compilers
reorder programs even more dramatically.  All of this reordering is meant to
be unobservable in single-threaded code, but is observable in multi-threaded code.
A formal attempt to understand the resulting mess is known
as a ``relaxed memory model.''  The relaxed memory models that have been
proposed to date either fail to address sequential composition, or overly
restrict processors and compilers.

To support sequential composition while targeing modern hardware, we propose adding families of predicate
transformers to the existing model of ``Pomsets with Preconditions,'' which
already supports parallel composition.  When composing $(S_1;S_2)$, the predicate
transformer used to validate the precondition of an event in $S_2$ is chosen
based on the semantic dependencies from events in $S_1$ into this event in $S_2$.  Our model
retains the good properties of the prior work, including efficient
implementation on Arm8, support for compiler optimizations, support for
logics that prove the absence of thin-air behaviors, and a local data race
freedom theorem.
